Generating Random Logic Programs Using Constraint Programming

### 2020-05-05 paper2/paper.tex
Improve the explanation of acyclicity

@@

  \keywords{Probabilistic logic programming \and Logic programming \and
    Predicate independence.}
\end{abstract}

% TODO (for later)
% 1) First keyword should be the category that I'm submitting for
% 2) Use American English
% 3) Avoid the Definition environment (to save space)
% 4) Make sure that figures and algorithms are placed in good positions

@@

@@

In this work, we introduce a constraint model for generating random logic
programs according to a number of user-specified parameters on the structure of
the program. In fact, the same model can generate both probabilistic programs
directly in the syntax of ProbLog \cite{DBLP:conf/ijcai/RaedtKT07} and
non-probabilistic Prolog programs. For generated probabilistic programs to be
valid, we use a custom constraint to detect negative cycles. A major advantage
of our constraint-based approach is that one can easily add additional
constraints to the model. To demonstrate that, we present a custom constraint
with propagation and entailment algorithms that can ensure predicate
independence. We also present a combinatorial argument for correctness, counting
the number of programs that the model produces for various parameter values.
Finally, we show how the model scales when tasked with producing more
complicated programs and identify the relationships between parameter values and
the empirical hardness of the program generation task.

@@

@@

The basic primitives of logic programs are \emph{constants}, \emph{(logic)
  variables}, and \emph{predicates}. Each predicate has an \emph{arity} that
defines the number of terms that it can be applied to. A \emph{term} is either a
variable or a constant, and an \emph{atom} is a predicate of arity $n$ applied
to $n$ terms. A \emph{formula} is a grammatically-valid expression that connects
atoms using conjunction ($\land$), disjunction ($\lor$), and negation ($\neg$).
A \emph{clause} is a pair of a \emph{head} (which is an atom) and a \emph{body}
(which is a formula). A \emph{(logic) program} is a multiset of clauses. Given a
program $\mathscr{P}$, a \emph{subprogram} $\mathscr{R}$ of $\mathscr{P}$ is a
subset of the clauses of $\mathscr{P}$ and is denoted by $\mathscr{R} \subseteq
\mathscr{P}$.

@@

We also define a measure of how complicated a body of a clause can become. As
each body is represented by a tree (see \cref{sec:bodies}), we set
$\maxNumNodes{} \ge 1$ to be the maximum number of nodes in the tree
representation of any clause. We also set $\maxNumClauses{}$ to be the maximum
number of clauses in a program. We must have that $\maxNumClauses{} \ge
|\predicates{}|$ because we require each predicate to have at least one clause
that defines it. The model supports eliminating either all cycles or just
negative cycles (see \cref{sec:cycles}) and enforcing predicate independence
(see \cref{sec:independence}), so a set of independent pairs of predicates is
another parameter. Since this model can generate probabilistic as well as
non-probabilistic programs, each clause is paired with a probability which is
randomly selected from a given multiset (i.e., our last parameter). For
generating non-probabilistic programs, one can set this list equal to $\{ 1 \}$.
Finally, we define $\tokens{} = \{ \neg, \land, \lor, \top \}$ as the set of
tokens that (together with atoms) form a clause. All decision variables of the
model can now be divided into $2 \times \maxNumClauses{}$ separate groups,
treating the body and the head of each clause separately. We say that the
variables are contained in two arrays: $\arrayt{bodies}{\maxNumClauses{}}{Body}$
and $\arrayt{heads}{\maxNumClauses{}}{Head}$. Since the order of the clauses
does not change the meaning of the program, we can also state our first
constraint:

\begin{constraint}
  Clauses are sorted.
\end{constraint}

Here and henceforth, the exact ordering is immaterial: we only impose an order
to eliminate permutation symmetries.

@@

\begin{definition}
  The \emph{head} of a clause is composed of a $\variable{predicate} \in
  \predicates \cup \{ \Box \}$, and
  $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}} \cup \{ \Box
  \}$.
\end{definition}

Here, we use $\Box$ to denote either a disabled clause that we choose not to use
or disabled arguments if the arity of the $\variable{predicate}$ is less than
$\maxArity{}$. The reason why we need a separate value for the latter (i.e., why
it is not enough to fix disabled arguments to a single already-existing value)
will become clear in \cref{sec:variable_symmetry}.

\begin{definition} \label{def:arity}
  The \variable{predicate}'s $\variable{arity} \in [0, \maxArity{}]$ can then be
  defined using the \variable{table} constraint as the arity of the
  $\variable{predicate}$ if $\variable{predicate} \in \predicates{}$, and zero
  otherwise.
\end{definition}

Having defined arity, we can now fix the superfluous arguments:

@@

\begin{constraint}
  Let $P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}$. Then
  \[
    \variable{nValues}(P) = \begin{cases}
      |\predicates{}| & \text{if } \variable{count}(\Box, P) = 0 \\
      |\predicates{}| + 1 & \text{otherwise,}
    \end{cases}
  \]
  where $\variable{nValues}(P)$ counts the number of unique values in $P$.
\end{constraint}

@@

As was briefly mentioned before, the body of a clause is represented by a tree.

% TODO: fix the formatting issue
\begin{definition}
  The \emph{body} of a clause has two parts. First, we have the
  $\arrayd{structure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ array that
  encodes the structure of the tree using the following two rules:
  $\variable{structure}[i] = i$ means that the $i$-th node is a root, and
  $\variable{structure}[i] = j$ (for $j \ne i$) means that the $i$-th node's
  parent is node $j$. The second part is the array
  $\arrayt{values}{\maxNumNodes{}}{Node}$ such that $\variable{values}[i]$ holds
  the value of the $i$-th node.
\end{definition}

@@

\begin{definition}
  For convenience, we also define $\variable{numNodes} \in \{ 1, \dots,
  \maxNumNodes{} \}$ to count the number of nodes in the main tree. We define it
  as
  \[
    \variable{numNodes} = \maxNumNodes{} - \variable{numTrees} + 1.
  \]
\end{definition}

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then
  \[
    \neg\mathsf{P}(X) \lor (\mathsf{Q}(X) \land \mathsf{P}(X))
  \]
  corresponds to the tree in \cref{fig:example_tree} and can be encoded as:
  \begin{alignat*}{8}
    \variable{structure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{values} &= [{\lor}, &&{\neg}, &&{\land}, \mathsf{P}(&&X),
    \mathsf{Q}(&&X), \mathsf{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
\end{example}

@@

Here, $\top$ is the value we use for the remaining one-node trees. The
elements of the $\variable{values}$ array are nodes:

\begin{definition} \label{def:node}
  A \emph{node} has a $\variable{name} \in \tokens{} \cup \predicates{}$ and
  $\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{} \cup \{ \Box
    \}}$. The node's $\variable{arity}$ can then be
  defined analogously to \cref{def:arity}.
\end{definition}

Furthermore, we can use \cref{constr:arity} to again disable the extra
arguments.

\begin{example}
  Let $\maxArity{} = 2$, $X \in \variables{}$, and let $\mathsf{P}$ be a
  predicate with arity 1. Then the node representing atom $\mathsf{P}(X)$ has:
  \begin{align*}
    \variable{name} &= \mathsf{P},\\
    \variable{arguments} &= [X, \Box],\\
    \variable{arity} &= 1.
  \end{align*}
\end{example}

@@

@@

\begin{definition}
  Let $N = \maxArity{} \times (\maxNumNodes{} + 1)$, and let
  $\variable{terms}[N] \in \constants{} \cup \variables{} \cup \{ \Box
  \}$ be a flattened array of all arguments in a particular clause.

  Then we can use a channeling constraint to define
  $\variable{occ}[|\constants{}| + |\variables{}| + 1]$ as an array of subsets
  of $\{ 0, \dots, N-1 \}$ such that for all $i = 0, \dots, N - 1$, and $t \in
  \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    i \in \variable{occ}[t] \quad \iff \quad
    \variable{terms}[i] = t.
  \]
\end{definition}

@@

Finally, we define an auxiliary set variable to act as a set of possible values
that $\variable{intros}$ can take:

\begin{definition}
  Let $\variable{potentials} \subseteq \{ 0, \dots, N \}$ be such that for $v
  \in \variables{}$, $\variable{intros}[v] \in \variable{potentials}$.
\end{definition}

Using this new variable, we can add a constraint saying that non-predicate nodes
in the tree representation of a clause cannot have variables as arguments:

@@

\section{Counting Programs} \label{sec:counting}

@@

@@

In this section, we define a notion of predicate independence as a way to
constrain the probability distributions defined by the generated programs. We
also describe efficient algorithms for propagation and entailment checking.

\begin{definition}
  Let $\mathscr{P}$ be a probabilistic logic program. Its \emph{predicate
    dependency graph} is a directed graph $G_{\mathscr{P}} = (V, E)$ with the
  set of nodes $V$ consisting of all predicates in $\mathscr{P}$. For any two
  different predicates $\mathsf{P}$ and $\mathsf{Q}$, we add an edge from
  $\mathsf{P}$ to $\mathsf{Q}$ if there is a clause in $\mathscr{P}$ with
  $\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say that
  the edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as the
  head and at least one instance of $\mathsf{P}$ at the body such that the path
  from the root to the $\mathsf{P}$ node in the tree representation of the
  clause passes through at least one negation node. Otherwise, it is
  \emph{positive}. We say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a
  \emph{negative cycle} if $G_{\mathscr{P}}$ has a cycle with at least one
  negative edge.
\end{definition}

Labelling the edges as positive/negative will be immaterial for predicate
independence, but the same graph will play a crucial role in negative cycle
detection in the next section.

\begin{definition}
  Let $\mathsf{P}$ be a predicate in a program $\mathscr{P}$. The set of
  \emph{dependencies} of $\mathsf{P}$ is the smallest set $D_{\mathsf{P}}$ such
  that $\mathsf{P} \in D_{\mathsf{P}}$, and, for every $\mathsf{Q} \in
  D_{\mathsf{P}}$, all direct predecessors of $\mathsf{Q}$ in $G_{\mathscr{P}}$
  are in $D_{\mathsf{P}}$.
\end{definition}

\begin{definition}
  Two predicates $\mathsf{P}$ and $\mathsf{Q}$ are \emph{independent} if
  $D_{\mathsf{P}} \cap D_{\mathsf{Q}} = \emptyset$.
\end{definition}

@@

\begin{figure}
  \centering
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
      \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
      \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
      \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
      \draw[gray] (and) -- (parent);
      \draw[gray] (and) -- (not);
      \draw[gray] (not) -- (mother);
    \end{tikzpicture}
    \captionof{figure}{A tree representation of the body of \eqref{eq:example_clause}}
    \label{fig:example_tree2}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}
    \captionof{figure}{The predicate dependency graph of the program in
      \cref{ex:program}. Positive edges are labelled with `$+$', and negative
      edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
\end{figure}

@@

\begin{example} \label{example:independence}
  Consider this partially determined (fragment of a) program:
  \begin{align*}
    \Box(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z),\\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y),
  \end{align*}
  where $\Box$ indicates an unknown predicate with domain
  \[
    D_\Box = \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent},
    \mathsf{sibling} \}.
  \]
  The predicate dependency graph without positivity/negativity (as defined in
  \cref{def:adjacency_matrix}) is represented in
  \cref{fig:dependencies_matrix,fig:dependencies2}.

  % TODO: labels on the inside (check if they work)
  \begin{figure}
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[The adjacency matrix defined using \cref{def:adjacency_matrix} for
      \cref{example:independence}\label{fig:dependencies_matrix}]{
        $\begin{blockarray}{lcccc}
          \begin{block}{l!{\quad}(cccc)}
            \mathsf{father} & 0 & 0 & 0 & 0 \\
            \mathsf{mother} & 1 & 0 & 0 & 0 \\
            \mathsf{parent} & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
            \mathsf{sibling} & 0 & 0 & 0 & 0 \\
          \end{block}
        \end{blockarray}$
      }
    \end{minipage}
    \hfill
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[The predicate dependency graph that corresponds to
      \cref{fig:dependencies_matrix}. Dashed edges are undetermined---they may or
      may not exist.\label{fig:dependencies2}]{
        \centering
        \makebox[\linewidth]{
          \begin{tikzpicture}
            \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
            \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
            \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
            \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
            \draw[-{Stealth}] (parent) edge (father);
            \draw[-{Stealth}] (mother) edge (father);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
          \end{tikzpicture}
        }
      }
    \end{minipage}
    \caption{TODO}
  \end{figure}

@@

\section{Negative Cycles} \label{sec:cycles}
% although some work has been done in overcoming this limitation
% \cite{DBLP:journals/ijar/BuchmanP17}

Having no negative cycles in the predicate dependency graph is a requirement of
ProbLog that makes the program well-defined \cite{kimmig2009trading}. Ideally,
we would like to design a constraint for negative cycles similar to the
constraint for independence in the previous section. However, the difficulty
with creating a propagation algorithm for negative cycles is that there seems to
be no good way to extend \cref{def:adjacency_matrix} so that the adjacency
matrix captures positivity/negativity. Thus, we settle for an entailment
algorithm with no propagation.

@@

@@

\bibliographystyle{splncs04}
\bibliography{bibliography}

>>>>

% TODO (for later)
% 1) First keyword should be the category that I'm submitting for (add a new one)
% 2) Use American English
% 3) Avoid the Definition environment (to save space)
% 4) Make sure that figures and algorithms are placed in good positions
% 5) Check the grammar

% Empirical tasks:
% 1) make the CP model faster: apply the redundant constraints to predicates
% (not just variables). Check if everything works out.
% 2) motivate either for benchmarking or by comparing parameter learning with
% ProbFOIL
% New things I could talk about:
% 1) entailment for conditional independence
% 2) the 'required formula' part of the model

@@

  \keywords{Probabilistic logic programming \and Statistical relational
    artificial intelligence.}
\end{abstract}

@@

% TODO: Review: I’m not convinced that paragraph 2 follows from paragraph 1. Has
% generation of random SAT problems really solved the problem of establishing
% that one inference algorithm is better than another? I thought it led to the
% opposite–a discovery that randomly generated instances are a different kind of
% problem than real-world benchmarks. For example, a recent SAT competition
% shows now overlap in the top three solvers for the random vs non-random
% tracks.

@@

In this work, we introduce a constraint model for generating random logic
programs according to a number of user-specified parameters on the structure of
the program. In fact, the same model can generate both probabilistic programs
in the syntax of ProbLog \cite{DBLP:conf/ijcai/RaedtKT07} and
non-probabilistic Prolog programs. A major advantage of our constraint-based
approach is that one can easily add additional constraints to the model. To
demonstrate that, we present a custom constraint with propagation and entailment
algorithms that can ensure predicate independence. We also present a
combinatorial argument for correctness, counting the number of programs that the
model produces for various parameter values. Finally, we show how the model
scales when tasked with producing more complicated programs.

A common restriction put on both logic programs and probabilistic logic programs
in order to ensure that there is a unique answer to every query is known as
\emph{stratification}
\cite{DBLP:journals/jlp/BalbinPRM91,DBLP:conf/padl/MantadelisR17}. We extend
this idea to programs that have arbitrary well-formed formulas in clause bodies
(as ProbLog supports this) and describe a custom constraint that ensures
stratification by checking for negative cycles in the predicate dependency
graph.

% TODO: Review: I would have appreciated a definition of 
% predicate independence instead of just a forward pointer to sections 7 and 8.

@@

% TODO: Answer to a review: The main benefit of a constraint-based approach is
% the ability to add additional constraints as needed, and to do that
% efficiently (compared to generate-and-test approaches). Specifically, we can
% regulate what parts of the program must be independent, require certain
% formulas to be included, etc.

% TODO: Review: The paper fails to provide explanation of critical parts: for
% example, how is a set of constraints going to be turned into a random logic
% program? We don’t find out until section 9.

@@

The basic primitives of logic programs are \emph{constants}, \emph{(logic)
  variables}, and \emph{predicates}. Each predicate has an \emph{arity} that
defines the number of terms that it can be applied to. A \emph{term} is either a
variable or a constant, and an \emph{atom} is a predicate of arity $n$ applied
to $n$ terms. A \emph{formula} is any well-formed expression that connects
atoms using conjunction ($\land$), disjunction ($\lor$), and negation ($\neg$).
A \emph{clause} is a pair of a \emph{head} (which is an atom) and a \emph{body}
(which is a formula). A \emph{(logic) program} is a multiset of clauses. Given a
program $\mathscr{P}$, a \emph{subprogram} $\mathscr{R}$ of $\mathscr{P}$ is a
subset of the clauses of $\mathscr{P}$ and is denoted by $\mathscr{R} \subseteq
\mathscr{P}$.
% TODO: Review: The use of disjunction and arbitrary negation in the body is
% also uncommon

@@

We also define a measure of how complicated a body of a clause can become. As
each body is represented by a tree (see \cref{sec:bodies}), we set
$\maxNumNodes{} \ge 1$ to be the maximum number of nodes in the tree
representation of any clause. We also set $\maxNumClauses{}$ to be the maximum
number of clauses in a program. We must have that $\maxNumClauses{} \ge
|\predicates{}|$ because we require each predicate to have at least one clause
that defines it.

% TODO: make the first sentence clear
The model supports enforcing predicate independence (see
\cref{sec:independence}), so a set of independent pairs of predicates is another
parameter. Since this model can generate probabilistic as well as
non-probabilistic programs, each clause is paired with a probability which is
randomly selected from a given multiset (i.e., our last parameter). For
generating non-probabilistic programs, one can set this list equal to $\{ 1 \}$.
Finally, we define $\tokens{} = \{ \neg, \land, \lor, \top \}$ as the set of
tokens that (together with atoms) form a clause. All decision variables of the
model can now be divided into $2 \times \maxNumClauses{}$ separate groups,
treating the body and the head of each clause separately. We say that the
variables are contained in two arrays: $\arrayt{bodies}{\maxNumClauses{}}{Body}$
and $\arrayt{heads}{\maxNumClauses{}}{Head}$. Since the order of the clauses
does not change the meaning of the program, we can also state our first
constraint:

\begin{constraint}
  Clauses are sorted according to a lexicographic increasing order\footnote{The
    exact order is immaterial: we only impose an order to eliminate duplicates
    and permutation symmetries.}.
\end{constraint}

@@

We define the \emph{head} of a clause as a $\variable{predicate} \in \predicates
\cup \{ \Box \}$ and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup
  \variables{}} \cup \{ \Box \}$. Here, we use $\Box$ to denote either a
disabled clause that we choose not to use or disabled arguments if the arity of
the $\variable{predicate}$ is less than $\maxArity{}$. The reason why we need a
separate value for the latter (i.e., why it is not enough to fix disabled
arguments to a single already-existing value) will become clear in
\cref{sec:variable_symmetry}.

This $\variable{predicate}$ variable has a corresponding arity that depends on
the value of $\variable{predicate}$. We can define $\variable{arity} \in [0,
\maxArity{}]$ as the arity of the $\variable{predicate}$ if
$\variable{predicate} \in \predicates{}$ and zero otherwise using the table
constraint \cite{DBLP:conf/cpaior/MairyDL15}. This constraint uses a set of
pairs of the form $(p, a)$, where $p$ ranges over all possible values of
$\variable{predicate}$, and $a$ is either the arity of predicate $p$ or zero.
Having defined arity, we can now fix the superfluous arguments:

@@

\begin{constraint}
  Let $P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}$ be a
  multiset. Then
  \[
    \variable{nValues}(P) = \begin{cases}
      |\predicates{}| & \text{if } \variable{count}(\Box, P) = 0 \\
      |\predicates{}| + 1 & \text{otherwise,}
    \end{cases}
  \]
  where $\variable{nValues}(P)$ counts the number of unique values in $P$, and
  $\variable{count}(\Box, P)$ counts how many times $\Box$ appears in $P$.
\end{constraint}

@@

As was briefly mentioned before, the \emph{body} of a clause is represented by a
tree. It has two parts. First, there is the
$\arrayd{structure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ array that encodes
the structure of the tree using the following two rules:
$\variable{structure}[i] = i$ means that the $i$-th node is a root, and
$\variable{structure}[i] = j$ (for $j \ne i$) means that the $i$-th node's
parent is node $j$. The second part is the array
$\arrayt{values}{\maxNumNodes{}}{Node}$ such that $\variable{values}[i]$ holds
the value of the $i$-th node.
% TODO: explain what Node is.

@@

For convenience, we also define $\variable{numNodes} \in \{ 1, \dots,
\maxNumNodes{} \}$ to count the number of nodes in the main tree. We define it
as
\[
  \variable{numNodes} = \maxNumNodes{} - \variable{numTrees} + 1.
\]

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then $\neg\mathsf{P}(X) \lor (\mathsf{Q}(X) \land
  \mathsf{P}(X))$ corresponds to the tree in \cref{fig:example_tree} and can be
  encoded as:
  \begin{gather*}
    \begin{alignedat}{8}
      \variable{structure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
      \variable{values} &= [{\lor}, &&{\neg}, &&{\land}, \mathsf{P}(&&X),
      \mathsf{Q}(&&X), \mathsf{P}(&&X), &&\top, &&\top],
    \end{alignedat} \\
    \variable{numNodes} = 6, \quad \variable{numTrees} = 3.
  \end{gather*}
\end{example}

@@

Here, $\top$ is the value we use for the remaining one-node trees. The
elements of the $\variable{values}$ array are nodes. A \emph{node} has a
$\variable{name} \in \tokens{} \cup \predicates{}$ and
$\arrayd{arguments}{\maxArity{}}{\variables{} \cup \constants{} \cup \{ \Box
  \}}$. The node's $\variable{arity}$ can then be defined in the same way as in
\cref{sec:heads}. Furthermore, we can use \cref{constr:arity} to again disable
the extra arguments.

\begin{example}
  Let $\maxArity{} = 2$, $X \in \variables{}$, and let $\mathsf{P}$ be a
  predicate with arity 1. Then the node representing atom $\mathsf{P}(X)$ has:
    $\variable{name} = \mathsf{P}$, $\variable{arguments} = [X, \Box]$,
    $\variable{arity} = 1$.
\end{example}

@@

% TODO: Response to a review: Symmetry removal is important for two reasons.
% First, if we want to generate all programs that match a certain description,
% we would rather not generate the same program multiple times. Second, symmetry
% removal makes the constraint solving task much simpler by significantly
% reducing the search space that needs to be explored, often turning an
% intractable problem into an easy one.

@@

\begin{definition}
  Let $N = \maxArity{} \times (\maxNumNodes{} + 1)$, and let
  $\variable{terms}[N] \in \constants{} \cup \variables{} \cup \{ \Box
  \}$ be a flattened array of all arguments in a particular clause. Then we can
  use a channeling constraint to define $\variable{occ}[|\constants{}| +
  |\variables{}| + 1]$ as an array of subsets of $\{ 0, \dots, N-1 \}$ such that
  for all $i = 0, \dots, N - 1$, and $t \in \constants{} \cup \variables{} \cup
  \{ \Box \}$,
  \[
    i \in \variable{occ}[t] \quad \iff \quad
    \variable{terms}[i] = t.
  \]
\end{definition}

@@

Finally, we define an auxiliary set variable to act as a set of possible values
that $\variable{intros}$ can take. Let $\variable{potentials} \subseteq \{ 0,
\dots, N \}$ be such that for $v \in \variables{}$, $\variable{intros}[v] \in
\variable{potentials}$. Using this new variable, we can add a constraint saying
that non-predicate nodes in the tree representation of a clause cannot have
variables as arguments:

@@

\section{Counting Programs} \label{sec:counting}
% TODO: Does this section still work if I don't allow duplicates?

@@

% TODO: Review: Finally, regarding correctness of the proposal, I don´t
% understand "the combinatorial argument for correctness" result, how does the
% provided analysis show correctness relative to the properties? I was expecting
% a proof that would show the relationship between the formalized constraint and
% the desired property, to chow for instance that if such constraint is taken
% into account in the construction of the program then it is impossible that the
% program does not satisfy the property. Is there something I do not understand
% from this? I find the proposal lacking in this terms, not finding the
% arguments for validity strong enough, unless I misunderstood the analysis.
% Answer: 
% Regarding correctness, we decided against including proofs that a constraint for
% a property ensures the property because the discussed properties are all
% syntactic, and the constraint model itself is a representation of the same
% syntax so the proofs would not be particularly revealing. We can nonetheless add
% them if necessary.
%
% Instead, Section 6 shows correctness in a sense that if two methods generate the
% same numbers (i.e., numbers of programs with certain properties and parameters),
% then they describe the same thing, and we gain confidence that the constraint
% description (which matches the combinatorial description) accurately matches our
% intentions. This method helped us debug many aspects of the model and can be
% particularly revealing when both the model and the combinatorial argument become
% more intricate.

@@

% TODO: Review: Section 7 defined predicate (in)dependence. It is done using the
% constraint model, but given a logic program, it could just as well be
% extracted from the logic program (construct body tree from each clause).
% Answer: Motivate why we need this.

% TODO: Answer to a review: All constraints have propagation and entailment
% procedures behind them. When designing a new constraint, unless it can be
% composed of already-existing constraints, designing these algorithms is, in
% fact, completely in the spirit of constraint programming.

In this section, we define a notion of predicate independence as a way to
constrain the probability distributions defined by the generated programs. We
also describe efficient algorithms for propagation and entailment checking.
Let $\mathscr{P}$ be a probabilistic logic program. Its \emph{predicate
  dependency graph} is a directed graph $G_{\mathscr{P}} = (V, E)$ with the set
of nodes $V$ consisting of all predicates in $\mathscr{P}$. For any two
different predicates $\mathsf{P}$ and $\mathsf{Q}$, we add an edge from
$\mathsf{P}$ to $\mathsf{Q}$ if there is a clause in $\mathscr{P}$ with
$\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say that the
edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as the head
and at least one instance of $\mathsf{P}$ at the body such that the path from
the root to the $\mathsf{P}$ node in the tree representation of the clause
passes through at least one negation node. Otherwise, it is \emph{positive}. We
say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a \emph{negative cycle} if
$G_{\mathscr{P}}$ has a cycle with at least one negative edge. Labelling the
edges as positive/negative will be immaterial for predicate independence, but
the same graph will play a crucial role in designing a constraint for
stratification in the next section. In fact, a program $\mathscr{P}$ is
\emph{stratified} if $G_{\mathscr{P}}$ has no negative cycles.

\begin{definition}
  Let $\mathsf{P}$ be a predicate in a program $\mathscr{P}$. The set of
  \emph{dependencies} of $\mathsf{P}$ is the smallest set $D_{\mathsf{P}}$ such
  that $\mathsf{P} \in D_{\mathsf{P}}$, and, for every $\mathsf{Q} \in
  D_{\mathsf{P}}$, all direct predecessors of $\mathsf{Q}$ in $G_{\mathscr{P}}$
  are in $D_{\mathsf{P}}$. Two predicates $\mathsf{P}$ and $\mathsf{Q}$ are
  \emph{independent} if $D_{\mathsf{P}} \cap D_{\mathsf{Q}} = \emptyset$.
\end{definition}

@@

\begin{figure}[t]
  \centering
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
      \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
      \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
      \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
      \draw[gray] (and) -- (parent);
      \draw[gray] (and) -- (not);
      \draw[gray] (not) -- (mother);
    \end{tikzpicture}
    \captionof{figure}{A tree representation of the body of \eqref{eq:example_clause}}
    \label{fig:example_tree2}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}
    \captionof{figure}{The predicate dependency graph of the program in
      \cref{ex:program}. Positive edges are labelled with `$+$', and negative
      edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
\end{figure}

@@

\begin{example} \label{example:independence}
  Consider this partially determined (fragment of a) program:
  \begin{align*}
    \Box(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z),\\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y),
  \end{align*}
  where $\Box$ indicates an unknown predicate with domain
  \[
    D_\Box = \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent},
    \mathsf{sibling} \}.
  \]
  The predicate dependency graph without positivity/negativity (as defined in
  \cref{def:adjacency_matrix}) is represented in \cref{fig:example}.

  \begin{figure}
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[The adjacency matrix of the graph. The boxed value is the
      decision variable that will be propagated by
      \cref{alg:independence_propagation}.\label{fig:dependencies_matrix}]{
        $\begin{blockarray}{lcccc}
          \begin{block}{l!{\quad}(cccc)}
            \mathsf{father} & 0 & 0 & 0 & 0 \\
            \mathsf{mother} & 1 & 0 & 0 & 0 \\
            \mathsf{parent} & 1 & \fbox{\{ 0, 1 \}} & \{ 0, 1 \} & \{ 0, 1 \} \\
            \mathsf{sibling} & 0 & 0 & 0 & 0 \\
          \end{block}
        \end{blockarray}$
      }
    \end{minipage}
    \hfill
    \begin{minipage}{.49\textwidth}
      \centering
      \subfloat[A drawing of the graph. Dashed edges are undetermined---they may
      or may not exist.\label{fig:dependencies2}]{
        \centering
        \makebox[\linewidth]{
          \begin{tikzpicture}
            \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
            \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
            \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
            \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
            \draw[-{Stealth}] (parent) edge (father);
            \draw[-{Stealth}] (mother) edge (father);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (sibling);
            \draw[-{Stealth},dashed,darkgray] (parent) edge (mother);
          \end{tikzpicture}
        }
      }
    \end{minipage}
    \caption{The predicate dependency graph of \cref{example:independence}}
    \label{fig:example}
  \end{figure}

@@

\section{Negative Cycles} \label{sec:cycles}

Ideally, we would like to design a constraint for negative cycles similar to the
constraint for independence in the previous section. However, the difficulty
with creating a propagation algorithm for negative cycles is that there seems to
be no good way to extend \cref{def:adjacency_matrix} so that the adjacency
matrix captures positivity/negativity. Thus, we settle for an entailment
algorithm with no propagation.

@@

% TODO: Review: From section 9 it seems that this language has a very special
% handcrafted (heuristic) solution procedure that is not guaranteed to find a
% solution if one exists.
% Answer: Adding a restart policy helps with efficiency when generating larger
% programs, but if we want to find all possible programs, restarts can be
% disabled (and this is likely to only be relevant with low parameter values,
% where the number of possible programs is in thousands rather than millions).
% Note that combinations of parameter values that have no solutions (or are
% unlikely to have solutions) are unlikely to occur in practice (i.e., be of
% interest).

@@

\bibliographystyle{splncs04}
\bibliography{paper}
